A Simple Type Theory for Category Theory

### 2022-02-09 paper/lipics-v2021-sample-article.tex
kan extensions

\subsection{(co)-Limits and Kan Extensions}

As a further example of the synthetic approach to universal
properties, we consider \emph{Kan extensions}, which are a
generalization of limits, and their dual \emph{Kan lifts}, which
generalize co-limits.

The right Kan extension of $F$ along $J$ is the ``best'' (maximal) way to
fill in the following square:
\[% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAGEQBfU9TXfIRQBGclVqMWbACLdeIDNjwEiZYePrNWiDgHJu4mFADm8IqABmAJwgBbJGRA4ISAEzVNUnQCk5lm-aIjs5IohJabABiBlxAA
\begin{tikzcd}
C \arrow[d, "J"] \arrow[r, "F"] & D \\
C'                              &  
\end{tikzcd}\]
I.e. a functor $Ran_J F : C' \to D$ with a natural transformation
$Ran_J F \Rightarrow F$ that is universal something something.

If $D$ is complete, there is a formula using ends for this. Using hom
of profunctors, we can directly adapt this to define a
\emph{profunctor} that is a universal filler. We can then say that a
right Kan extension is precisely a functor that represents this
profunctor.

\begin{definition}
  Let $C,C',D$ be categories and let $F : \textrm{Functor}\,C\,D$ and
  $J: \textrm{Functor}\,C'\,C$ be functors. Then a right Kan extension
  is a functor $\textrm{Ran}_JF :\varr {\cat C'}{\cat D}$ that
  right-represents the following profunctor:
  \[ \]

  Then the right Kan
  extension specification profunctor is a profunctor
  $\textrm{RanSpec}_J\,F : \textrm{Prof}\,D\,C$ defined as

  \[ \textrm{RanSpec}_J\,F = \textrm{Hom}(C')[;J] \triangleright \textrm{Hom}(D)[;F] \]

  And a right Kan extension is a functor
  $\textrm{Ran}_JF:\textrm{Functor}\,C'\,D$ that is a
  right-representation of $\textrm{RanSpec}_J F$.
\end{definition}

Writing this out more pointfully, we get that the hom set
$\harrapp{D_1} {\beta}{\textrm{Ran}_JF\, \alpha}$ is an equalizer of
the following spans:
\[ \homrallXYtoZ {\alpha':C'} {\harrapp {C_1'} {\alpha'}{J_0\alpha}}{\harrapp {D_1} {\beta} {F_0 \alpha'}}\]

If we have a unit and therefore unit category and functor, then we can
now define limits as a special case of Kan extensions.
\begin{definition}
  The limit of a functor $J : \textrm{Functor}I\,C$ is a right Kan
  extension $\textrm{Ran}_E J$ where $E$ is the unique functor $E :
  \textrm{Functor}I\,1$.
\end{definition}

Unraveling this definition, we get that $\harrapp{C_1} \alpha {\lim_{i} J(i)}$ is a coherent family (i.e., equalizer) of
\[  \homrallXYtoZ {i:I} {1} {\harrapp{C_1} \alpha {J_0(i)}}  \]
with the equalizing condition being that for any $p : \homrallXYtoZ
{i:I} {1} {\harrapp{C_1} \alpha {J_0(i)}}$ and $f : \harrapp{I_1} {i'}{i}$ and $g : \harrapp{1}{pt}{E_0 i'}$
\[ \homrappXtoYatZ {p} {(\appbinary {1_t} {pt} {\pendappXtoY {1_r} {pt}} {pt} {(\appbinary {1_t} {pt} g {E_0 i'} {(\appunary {E_1} {i'} f {i})} {i})} {i})} {i}
= \appbinary {C_t} {\alpha} {(\homrappXtoYatZ {p} g {i'})} {J_0 i'} {(\appbinary {C_t} {J_0 i'} {(\appunary {J_1} i f {i'})} {J_0 i} {\pendappXtoY {C_r} {J_0 i}} {J_0 i})} {J_0 i}
\]
We can greatly simplify using the $\eta$ principle for $1$ and
eliminating some reflexivities to reduce this to
\[ \homrappXtoYatZ {p} {()} {i} =
\appbinary {C_t} {\alpha} {(\homrappXtoYatZ {p} {()} {i'})} {J_0 i'} {{(\appunary {J_1} i f {i'})}} {J_0 i}
\]
which reads exactly as the commuting condition for a cone.

We can similarly define \emph{left} Kan extensions by a dual
formula. Left Kan extensions are the ``minimal'' solution to the same
filling problem as right Kan extensions. I.e., they have a universal
morphism out of them. They can be described by a coend formula, but in
our profunctor calculus, they are easiest to describe using a similar
formula to right Kan extensions, but using the contravariant hom of
profunctors.

\begin{definition}
  Let $C,C',D$ be categories and let $F : \textrm{Functor}\,C\,D$ and
  $J: \textrm{Functor}\,C'\,C$ be functors. Then the left Kan
  extension specification profunctor is a profunctor
  $\textrm{LanSpec}_J\,F : \textrm{Prof}\,C\,D$ defined as

  \[ \textrm{RanSpec}_J\,F = \textrm{Hom}(D)[F;] \triangleleft \textrm{Hom}(C')[J;]  \]
  And a left Kan extension is a functor $\textrm{Lan}_J F :
  \textrm{Functor}\,C'\,D$ that represents this profunctor on the
  left.
\end{definition}

Unraveling the definition we get the hom span $\harrapp {D_1}{\textrm{Ran}_JF {\alpha'}} \beta$ is an equalizer of
\[ \homlallXYtoZ \alpha {\harrapp {C_1'} {J_0 \alpha} {\alpha'}} {\harrapp{D_1} {F_0 \alpha} \beta}\]

\begin{definition}
  The colimit of a functor $J : \textrm{Functor}I\,C$ is a left Kan
  extension $\textrm{Lan}_E J$ where $E$ is the unique functor $E :
  \textrm{Functor}I\,1$.
\end{definition}

We can similarly unravel this definition to see that this looks
essentially like the classical definition of colimits as universal
cocones.

\subsection{Right (Left) Adjoints preserve (co-)Limits}

Let's prove that classic theorem!

\begin{theorem}
  Let $\alpha : \cat C \vdash A \cat D$ be an arbitrary functor and
  $\beta : \cat D \vdash R : \cat E$ be a functor with a left adjoint
  $L$. Then, if the limit $\lim_\alpha A$ exists, then $\lim_\alpha
  R[A/\beta]$ also exists and is
  \[ \lim_\alpha R[A/\beta] = R[\lim_\alpha A] \]
\end{theorem}
\begin{proof}
  We need to give an isomorphism
  \[ \gamma \to R[\lim_\alpha A/\beta] \cong PRan_{()}{\alpha. R[A/\beta]} = \forall \alpha. (\gamma \to R[A/\beta]) \triangleleft (() \to_1 ()) \]
  By assumption we have an isomorphism
  \[ l : \beta \to \lim_\alpha A \cong PRan_{()}{\alpha. A} = \forall \alpha. (\beta \to A[\alpha]) \triangleleft (()\to_1 ())\]
  TODO.
\end{proof}

%% \subsection{Additive Facts}

%% I think the following isomorphisms should hold, which connect additive
%% connectives on objects with the analogous connective on profunctors:
%% \[ () \to_1 () \cong () \]
%% \[ (\pi_1 \alpha, \pi_2 \alpha) \to_{\cat C \times \cat D} (\pi_1 \beta, \pi_2 \beta) \cong
%% ((\pi_1 \alpha) \to_{\cat C} (\pi_1 \beta)) \times ((\pi_2 \alpha) \to_{\cat C} (\pi_2 \beta)) \]

>>>>

\subsection{Kan Extensions/Lifts}

As a further example of the synthetic approach to universal
properties, we consider right and left \emph{Kan extensions}, which
are generalizations of limits and colimits, respectively.

The right Kan extension of $F$ along $J$ is the ``best'' (maximal) way to
fill in the following square:
\[% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAGEQBfU9TXfIRQBGclVqMWbACLdeIDNjwEiZYePrNWiDgHJu4mFADm8IqABmAJwgBbJGRA4ISAEzVNUnQCk5lm-aIjs5IohJabABiBlxAA
\begin{tikzcd}
\cat C \arrow[d, "J"] \arrow[r, "F"] & \cat D \\
\cat C'                              &  
\end{tikzcd}\]
I.e. a functor $\textrm{Ran}_J F : \cat C' \to \cat D$ with a
universal natural transformation $\textrm{Ran}_J F \circ J \Rightarrow
F$. A left kan extension $\textrm{Lan}_J F$ is the same but with a
natural transformation $F \Rightarrow \textrm{Lan}_J F \circ J $.

We can express the concept in SCT as representing a certain profunctor
constructed using the profunctor hom:
\begin{definition}[Kan Extensions]
  Given functors $F : \varr {\cat C}{\cat D}$ and $J : \varr {\cat
    C}{\cat C'}$, a right Kan extension is a functor $\textrm{Ran}_J F
  : \varr {\cat C'}{\cat D} $that right-represents the profunctor
  \[ \hlambda {\beta:\cat D}{\alpha': \cat C} \homrallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{\alpha'}{J\alpha})} {\punitinXfromYtoZ {\cat D}{\beta}{F \alpha}} \]

  A left Kan extensions is a functor $\textrm{Lan}_J F : \varr {\cat
    C'}{\cat D}$ is a functor satisfying the dual property: it
  \emph{left}-represents the profunctor
  \[\hlambda {\alpha':\cat C}{\beta:\cat D} \homlallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{J\alpha}{\alpha'})} {\punitinXfromYtoZ {\cat D}{F \alpha}{\beta}}
  \]
\end{definition}

The limit of a functor $F : \varr{\cat C}{\cat D}$ is a special case
of a Kan extension where $J : \varr{\cat C}{1}$ is the unique functor
to the terminal category. Similarly, the colimit of $F$ is the left
Kan extension with respect to the same $J$.
%
Specializing to this case, the representability for limits means:
\[ \punitinXfromYtoZ {\cat D}{\beta}{\textrm{lim} F} \cong \homrallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat 1}{\alpha'}{()})}{\punitinXfromYtoZ {\cat D}{\beta}{F \alpha}} \]
And if we view a morphism in $\cat 1$ as being trivial, then this
essentially states that mapping $\beta$ to $\textrm{lim} F$ is the
same thing as constructing a cone: for all $\alpha :\cat C$, a
morphism from $\beta$ to $F \alpha$, naturally in $\alpha$.

The usual formulation of Kan extensions is in terms of a universal
morphism.
%
We can reproduce this formulation using our established theory of
equivalences for homs of profunctors.
\begin{lemma}
  One half of the representability isomorphism for a right Kan
  extension $\textrm{Ran}_J F$ is determined by a natural
  transformation
  \[ \pendallXdotY \alpha \punitinXfromYtoZ {\cat D}{(\textrm{Ran}_JF)(J\alpha)}{F\alpha} \]

  Dually, one half of the representability isomorphism for a left Kan
  extension $\textrm{Lan}_J F$ is determined by a natural
  transformation
  \[ \pendallXdotY \alpha \punitinXfromYtoZ {\cat D}{F\alpha}{(\textrm{Lan}_JF)(J\alpha)} \]
\end{lemma}
\begin{proof}
  By \cref{lem:unit-repr}, one half of representability is determined by a natural element
  \[ \pendallXdotY {\alpha'} \homrallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{\alpha'}{J\alpha})} {\punitinXfromYtoZ {\cat D}{(\textrm{Ran}_J F)\alpha'}{F \alpha}} \]
  which, by \cref{lem:fubini} is isomorphic to
  \[
  \pendallXdotY {\alpha} \homlallXYtoZ {\alpha'}{(\punitinXfromYtoZ {\cat C'}{\alpha'}{J\alpha})}{\punitinXfromYtoZ {\cat D}{(\textrm{Ran}_J F)\alpha'}{F \alpha}}
  \]
  which, by \cref{lem:yo-and-co-yo}, is isomorphic to 
  \[ \pendallXdotY \alpha \punitinXfromYtoZ {\cat D}{(\textrm{Ran}_JF)(J\alpha)}{F\alpha} \]
  by taking $P = \hlambda {\alpha'}{\alpha} {\punitinXfromYtoZ {\cat D}{(\textrm{Ran}_J F)\alpha'}{F \alpha}}$.

  The proof for left Kan extensions follows by \cref{lem:duality}.
\end{proof}

With the definition based on representability, we can easily prove the
generalization of the famous theorem, that right adjoints preserve not
just limits but all right Kan extensions

\begin{theorem}[Right/Left Adjoints preserve Right/Left Kan Extensions]
  Let $K : \varr{\cat C}{\cat D}$ and $J : \varr{\cat C}{\cat C'}$.

  If the right Kan extension $\textrm{Ran}_J K$ exists, and if $G :
  \varr {\cat D}{\cat D'}$ has a left adjoint $F : \varr{\cat D'}{\cat
    D}$, then the right Kan extension $\textrm{Ran}_J G\circ K$ exists
  and is given by $G \circ \textrm{Ran}_J K$.

  Similarly, if the left Kan extension $\textrm{Lan}_J K$ exists and
  if $F : \varr{\cat D}{\cat D'}$ has a right adjoint $G : \varr{\cat
    D'}{\cat D}$, then the left Kan extension $\textrm{Lan}_J F\circ
  K$ exists and is given by $F \circ \textrm{Ran}_J K$.
\end{theorem}
\begin{proof}
  By the composition of the following natural isomorphisms:
  \begin{align*}
    \punitinXfromYtoZ {\cat D'} {\beta'} {G ((\textrm{Ran}_J K) \alpha')} &\cong \punitinXfromYtoZ {\cat D}{F \beta'} {(\textrm{Ran}_J K) \alpha'} \tag{Adjunction}\\
    &\cong \homrallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{\alpha'}{J\alpha})} {\punitinXfromYtoZ {\cat D}{F \beta'}{K \alpha}} \tag{Ran assumption, \cref{lem:natural-restriction}}\\
    &\cong \homrallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{\alpha'}{J\alpha})} {\punitinXfromYtoZ {\cat D}{\beta'}{G(K \alpha)}} \tag{Adjunction}\\
  \end{align*}
\end{proof}
